Local Type Inference for Polarised System F with Existentials

### 2023-11-21 oopsla24/tex/05-proof.tex
wip

The proof of soundness and completeness result is done gradually
for all the subroutines of the algorithm,
following the 
structure of the algorithm 
(\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
bottom-up. 
\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, the correctness of normalization means that
    checking for equivalence of two types is the same as checking for equality of their normal forms.
    \begin{lemma*}[Normalization Correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we construct an intermediate relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

\subsection{Unification Constraint Merge}

\subsection{Unification}

\begin{lemma*}[Soundness of Unification]
    \label{lemma:unification-soundness}
    \hfill
    \begin{itemize}
        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
        $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
        such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.

        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
        $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
        $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
        $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Completeness of Unification]
    \label{lemma:unification-completeness}
    \hfill
    \begin{itemize}
        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
        suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
        then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.

        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
        suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
        then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
   \end{itemize}
\end{lemma*}

@@

\begin{lemma*}[Soundness of Anti-Unification]
    \label{lemma:au-soundness}
    \hfill
    \begin{itemize}
        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
        then 
        \begin{enumerate}
            \item $[[Γ ; Ξ ⊢ uQ]]$,
            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
            are anti-unification substitutions, and
            \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
        \end{enumerate}

        \item [$-$] Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
        then
        \begin{enumerate}
            \item $[[Γ ; Ξ ⊢ uM]]$,
            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
            are anti-unification substitutions, and
            \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
        \end{enumerate}
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Completeness of Anti-Unification]
    \label{lemma:au-completeness}
    \hfill
    \begin{itemize}
        \item [$+$] 
            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
            there exists $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uQ']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
                are anti-unification substitutions, and
                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
            \end{enumerate}

            Then the anti-unification algorithm terminates, that is there exists
            $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$

        \item [$-$] 
            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
            there exists $[[(Ξ', uM', aus'1, aus'2)]]$ such that
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uM']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
                are anti-unification substitutions, and
                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
            \end{enumerate}

            Then the anti-unification algorithm succeeds, that is 
            there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Initiality of Anti-Unification]
    \hfill
    \begin{itemize}
        \item [$+$] 
            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
            then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
            any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, i.e.
            if 
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uQ']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
            \end{enumerate}
            then there exists $[[ausr]]$ such that
            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
            Moreover, $[[ [ausr]β̂⁻]]$ can
            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
            $[[Γ]]$.
        \item [$-$] 
            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$, 
            then $[[(Ξ, uM, aus1, aus2)]]$ is more specific than
            any other sound anti-unifier $[[(Ξ', uM', aus'1, aus'2)]]$, i.e.
            if
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uM']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
            \end{enumerate}
            then there exists $[[ausr]]$ such that
            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
            Moreover, $[[ [ausr]β̂⁻]]$ can
            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
            $[[Γ]]$.
    \end{itemize}
\end{lemma*}

@@

\subsection{Positive Subtyping}

\begin{lemma*}[Soundness of the Positive Subtyping]
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any normalized $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of the Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
    Then for any $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{lemma*}

\subsection{Subtyping Constraint Merge}

\begin{lemma*}[Completeness of Constraint Merge]
    \label{lemma:merge-completeness}
    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$.
    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
    $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
\end{lemma*}

\begin{lemma*}[Soundness of Constraint Merge] 
    \label{lemma:merge-soundness}
    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$ 
    and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
    Then 
    \begin{enumerate}
        \item $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$,
        \item for any substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$, 
            $[[ Θ ⊢ uσ : SC ]]$
            implies $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
    \end{enumerate}
\end{lemma*}

\subsection{Negative Subtyping}

\begin{lemma*}[Soundness of Negative Subtyping]
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
    $[[Θ ⊢ SC : uv(uN)]]$ and 
    for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{lemma*}

@@

@@

\begin{lemma*}[Completeness of Singularity]
    \label{lemma:singularity-completeness}
    For a given $[[Θ ⊢ SC]]$,
    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
    on $[[Ξ]] \supseteq [[dom(SC)]]$.
    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ]]$ such that
    for any $[[Θ ⊢ uσ : Ξ]]$, $[[Θ ⊢ uσ : SC]]$ implies 
    $[[Θ ⊢ uσ ≈ uσ1 : Ξ]]$.
    Then 
    \begin{itemize}
        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
        \item $[[Ξ]] = [[dom(SC)]]$.
    \end{itemize} 
\end{lemma*}

\subsection{Type Inference}

>>>>

The formulation of soundness and completeness of \emph{application inference} is more complex.
Both of them assume that the input head type $[[uN]]$ is free from \emph{negative}
algorithmic variables (it is achieved by polarization invariants preserved by the inference rules).
The soundness states that the output of the algorithm---$[[uM]]$ and $[[SC]]$ is viable. 
Specifically, that the constraint set $[[SC]]$ provides a sufficient set of restrictions that 
a substitution $[[uσ]]$ must satisfy to ensure the \emph{declarative} inference of the output type $[[uM]]$, that is 
$[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$.

The application inference completeness means that if 
there exists a substitution $[[uσ]]$ and the resulting type $[[iM]]$
ensuring the declarative inference $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$ then the algorithm succeeds
and gives the most general result $[[uM0]]$ and $[[SC0]]$.
The property of `being the most general' is specified in pt. (\ref{point:mostGeneral}).
Intuitively, it means that for any other solution---substitution $[[uσ]]$ and the resulting type $[[iM]]$,
if it ensures the declarative inference, then $[[uσ]]$ can be extended in a $[[SC0]]$-complying way to equate 
$[[uM0]]$ with $[[iM]]$.

\footnotetext[1]{The presented properties hold, 
            but the actual inductive proof requires strengthening 
            of the statement and the corresponding theorem is more involved. 
            See the appendix for details.}

\begin{theorem*}[Soundness of Typing]
    \label{thm:soundness-typing}
    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
    \hfill
    \begin{itemize}
        \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$,
        \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$,
        \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$,
            assuming $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
            \emph{negative} algorithmic variables.
    \end{itemize}
\end{theorem*}

\begin{theorem*}[Completeness of Typing]
    \label{thm:completeness-typing}
    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
    \begin{itemize}
        \item [$+$] $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$,
        \item [$-$] $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$,
        \item [$\bullet$] If $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$
            where 
            \begin{enumerate*}
                \item $[[Γ ⊢ Θ]]$, 
                \item $[[Γ ⊢ iM]]$,
                \item $[[Γ; dom(Θ) ⊢ uN]]$ (free from \emph{negative} algorithmic variables), and
                \item $[[Θ ⊢ uσ : uv(uN)]]$,
            \end{enumerate*}
            then there exist $[[uM0]]$, $[[Θ0]]$, and $[[SC0]]$ such that
            \begin{enumerate}
                \item $[[ Γ; Φ; Θ ⊨ uN ● args ⇒> uM0 ⫤ Θ0; SC0 ]]$ and
                \item \label{point:mostGeneral} for any other $[[uσ]]$ and $[[iM]]$ 
                (where $[[Θ ⊢ uσ : uv(uN)]]$ and $[[Γ ⊢ iM]]$)
                    such that $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$, 
                    there exists $[[uσ']]$ such that 
                    \begin{enumerate*}
                        \item $[[Θ0 ⊢ uσ' : uv uN ∪ uv uM0]]$
                            and $[[Θ0 ⊢ uσ' : SC0]]$,
                        \item $[[Θ ⊢ uσ' ≈ uσ : uv uN]]$, and 
                        \item $[[Γ ⊢ [uσ']uM0 ≈ iM]]$.
                    \end{enumerate*}
            \end{enumerate}
    \end{itemize}
\end{theorem*}

The proof of soundness and completeness result is done gradually
for all the subroutines,
following the structure of the algorithm 
(\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
bottom-up. Next, we discuss the main of these results.

\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, the correctness of normalization means that
    checking for equivalence of two types is the same as checking for equality of their normal forms.
    \begin{lemma*}[Normalization Correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we introduce an \emph{intermediate} relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

% \subsection{Unification Constraint Merge}

% \subsection{Unification}

% \begin{lemma*}[Soundness of Unification]
%     \label{lemma:unification-soundness}
%     \hfill
%     \begin{itemize}
%         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
%         $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
%         if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
%         $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
%         such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.

%         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
%         $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
%         if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
%         $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
%         $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
%     \end{itemize}
% \end{lemma*}

% \begin{lemma*}[Completeness of Unification]
%     \label{lemma:unification-completeness}
%     \hfill
%     \begin{itemize}
%         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
%         $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
%         suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
%         then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.

%         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
%         $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
%         suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
%         then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
%    \end{itemize}
% \end{lemma*}

@@

The anti-unifier of the two types is the most specific pattern that matches against both of them. 
In our setting, the anti-unifiers are restricted further: first, the pattern might only contain placeholders at 
\emph{negative} positions (because eventually, the placeholders become variables abstracted by the existential quantifier);
second, the anti-unification is parametrized with a context $[[Γ]]$, in which the pattern instantiations must be well-formed.

This way, the correctness properties of the anti-unification algorithm are refined accordingly.
The soundness of anti-unification not only ensures that the resulting pattern matches with 
the input types, but also that the pattern instantiations are well-formed in the corresponding context, and 
that all the `placeholder' variables are negative. In turn, completeness states that if there exists a solution 
satisfying the soundness criteria, then the algorithm succeeds.

The correctness properties are formulated by the following lemmas. For brevity, we only provide the statements for 
the positive case, since the negative case is symmetric.

\begin{lemma*}[Soundness of (Positive) Anti-Unification]
    \label{lemma:au-soundness}
     Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
    if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
    then 
    \begin{enumerate*}
        \item $[[Γ ; Ξ ⊢ uQ]]$,
        \item $[[Γ ; · ⊢ aus1 : Ξ]]$ for $i \in \{1,2\}$
        are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
        \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}
\end{lemma*}

\begin{lemma*}[Completeness of (Positive) Anti-Unification]
    \label{lemma:au-completeness}
    Assuming that $[[iP1]]$ and $[[iP2]]$ are normalized terms well-formed in $[[Γ]]$
    and there exist  $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
    \begin{enumerate*}
        \item $[[Γ ; Ξ' ⊢ uQ']]$,
        \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
        are anti-unification substitutions, and
        \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}

    Then the anti-unification algorithm terminates, that is there exists
    $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$.
\end{lemma*}

The anti-unification is used to find the least upper bound (LUB).
To guarantee that the result of the LUB algorithm is indeed the least, 
we also need to show the corresponding property of anti-unification.
We prove that the anti-unifier that the algorithm provides 
is the most specific (or the most `detailed'). Specifically, 
it means that any other sound anti-unification solution 
can be `refined' to the result of the algorithm.  
The `refinement' is represented as an instantiation of the anti-unifier---a 
substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the placeholders $[[Ξ1]]$ with
types that themselves can contain placeholders from $[[Ξ2]]$.

\begin{lemma*}[Initiality of Anti-Unification]
        Assume that $[[iP1]]$ and $[[iP2]]$ are normalized types well-formed in $[[Γ]]$, 
        and the anti-unification algorithm succeeds: $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$. 
        Then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
        any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, \ie if
        \begin{enumerate*}
            \item $[[Γ ; Ξ' ⊢ uQ']]$,
            \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
            \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
        \end{enumerate*}
        then there exists a `refining' substitution $[[ausr]]$ such that
        $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
        % Moreover, $[[ [ausr]β̂⁻]]$ is uniquely determined by $[[ [aus'1]β̂⁻ ]]$, 
        % $[[ [aus'2]β̂⁻ ]]$, and $[[Γ]]$.
\end{lemma*}

To prove the correctness properties of the anti-unification algorithm,
one extra observation is essential. The algorithm relies on the fact that
in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
there are no two different placeholders $[[β̂⁻]]$
mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
This is important to guarantee that, for example, 
the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
$[[↓↑Bool → ↑Bool]]$ is $[[↓â⁻ → â⁻]]$, but not
(less specific) $[[↓â⁻ → b̂⁻]]$.

\begin{lemma*}[Uniqueness of Anti-unification Variable Names]
    Names of the anti-unification variables are uniquely defined by
    the types they are mapped to by the resulting substitutions. 
    Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
        then for any $[[β̂⁻]] \in [[Ξ]]$,
        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$.
\end{lemma*}

@@

\subsection{Subtyping}

    As for other properties, the correctness of subtyping means that
    the algorithm produces a valid result (soundness) 
    whenever it exists (completeness).
    In \cref{fig:subtyping-algo}, one can see that the positive and 
    negative subtyping relations are not \emph{mututally} recursive: 
    negative subtyping algorithm uses the positive subtyping, 
    but not vice versa. 
    Because of that, the inductive proofs of the \emph{positive} subtyping correctness are done independently.

    The soundness of positive subtyping states that
    the output constraint $[[SC]]$ provides a sufficient set of restrictions.
    In other words, any substitution satisfying $[[SC]]$, 
    ensures the desired declarative subtyping: $[[Γ ⊢ [uσ]uP ≥ iQ]]$.
    Notice that the soundness requires that only the left-hand side input 
    type ($[[uP]]$) is algorithmic. This is one of the invariants of the algorithm
    that significantly simplifies the unification and constraint resolution.

\begin{lemma*}[Soundness of the Positive Subtyping]
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any  $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    we have $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

    The completeness of subtyping says that if the substitution 
    ensuring the declarative subtyping exists, then the algorithm terminates.

\begin{lemma*}[Completeness of the Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
    Then if there exists $[[uσ]]$ such that $[[Θ ⊢ uσ : uv(uP)]]$ and $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    then the subtyping algorithm succeeds: $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$.
\end{lemma*}

After the correctness properties of the positive subtyping are established,
they are used to prove the correctness of the negative subtyping.
The soundness is formulated symmetrical to the positive case,
however, the completeness requires an additional invariant to be preserved.
The algorithmic input type $[[uN]]$ must be free from \emph{negative} algorithmic variables.
In particular, it ensures that the constraint restricting a \emph{negative} algorithmic
variable will never be generated, and thus, we do not need the 
Greatest Common Subtype procedure to resolve the constraints.

\begin{lemma*}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
\end{lemma*}

@@

The rule \ruleref{\ottdruleDTAppLetLabel} of the declarative inference
says that one can omit the type annotation of a let-binding if the inferred 
type of the bound application is unique. Accordingly, the algorithm,
that returns the constraint set, must be able to check that
the set of constraints has a single solution, and also to find this solution. 
This subroutine is called singularity, and its correctness is formulated as follows.

The soundness states that $[[SC singular with uσ]]$
implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
on the domain.  The completeness states that if all the 
$[[SC]]$-compliant substitutions are equivalent, then the singularity procedure succeeds.

@@

\begin{lemma*}[Completeness of Singularity]
    \label{lemma:singularity-completeness}
    For a given $[[Θ ⊢ SC]]$,
    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
    on $[[dom(SC)]]$.
    In other words, suppose that there exists $[[Θ ⊢ uσ1 : dom(SC)]]$ such that
    for any $[[Θ ⊢ uσ : dom(SC)]]$, $[[Θ ⊢ uσ : SC]]$ implies 
    $[[Θ ⊢ uσ ≈ uσ1 : dom(SC)]]$.
    Then\footnotemark[1] $[[SC singular with uσ0]]$ for some $[[uσ0]]$.
\end{lemma*}

\subsection{Typing}

Finally, we discuss the proofs of the theorems stated at the beginning of this section.

\paragraph{Determinicity}
    One of the properties that our proof relies on is the determinicity of the 
    typing algorithm: the output (the type, the set of restrictions, \etc) 
    is uniquely determined by the input (the term, the context, \etc).
    Determinicity is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinicity of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

\begin{lemma*}[Determinicity of the Typing Algorithm]
    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
    \begin{itemize}
        \item [$+$] If $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ v : iP']]$ then $[[iP = iP']]$.
        \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
        \item If $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ and 
            $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
            $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
    \end{itemize}
\end{lemma*}

\paragraph{Mutual Induction of Soundness and Completeness}

The soundness and completeness of type inference cannot be proven separately:
the inductive proof of one requires another, and vice versa. 
Soundness proof requires completeness in the case of 
\ruleref{\ottdruleATAppLetLabel},
The soundness proof can be considered as a transformation from the algorithmic 
inference tree to a declarative one. In the case of \ruleref{\ottdruleATAppLetLabel},
the declarative rule requires that the (declaratively) inferred type of the application
is unique.  To show that, we take an arbitrary application typing tree and apply the typing
\emph{completeness}. Then the soundness of singularity
